---
title: "Introduction to geoChronR"
author: "Nick McKay"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to geoChronR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

### Welcome to geoChronR!

##Workflow 

Almost always, your geoChronR work flow will start with loading in a valid LiPD file. So you'll want both the geoChronR and lipdR libraries

```{r}
library(lipdR)
library(geoChronR)
```


In this case, we're going to load in a marine sediment core called ODP1098B:

```{r}
M = readLipd("ODP1098B13.lpd")
```

At this point - you might use `M = runBacon(M)` or `M = runBchron(M)`. In this case, a BChron model has already been stored in the LiPD file. We can plot the chron model and data like this:

```{r}
plotChron(M,model.num = 1)
```



## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
