---
title: "Aventures in Paleo Spectral Analysis - synthetic edition"
output: html_spectral
---
This little notebook is meant to explore various packages, namely astrochron and tuts, and how they can help identify harmonics in paleoclimate timeseries. Note that astrochron assumes evenly sampled, time-certain data (what were they smoking, exactly?), so we start with a synthetic example.

First install & load astrochron:

```{r}
#install.packages("astrochron")
library(astrochron)
```

Then generate example series with periods of 400 ka, 100 ka, 40 ka and 20 ka, and tiny bit of noise to make things easy at first.
```{r}
ex = astrochron::cycles(freqs=c(1/400,1/100,1/40,1/20),start=1,end=1000,dt=5,noisevar=0.1)
#noise = ar1(npts=200,dt=5,sd=0.5)
# ex[2] = ex[2] + noise[2]
```

Now apply the Multi-taper method, with the Mann & Lees estimate of the red noise background. Reduce the padding factor, which is 5 by default (seems like an awful lot)
```{r}
pl(1, title="mtmML96")
mtmML96.out <- astrochron::mtmML96(ex,tbw=3,padfac = 1, output =1, pl = 1)
```

Compare to test against a conventional AR(1) noise test

```{r}
pl(1,title="mtm")
astrochron::mtm(ex,padfac=1,ar1=TRUE)
```

compare to analysis with LOWSPEC (Meyers, 2012)
```{r}
pl(1, title="lowspec")
spec_low = astrochron::lowspec(ex,output=1)
```

oh dear, so many cycles breach the bakground! How exciting! Of course, this was all pretty dumb, because we weren't taking into account test multiplicity. (also: frequentist. boooh!)
Now let's use the multitest routine in astrochron. To make things easier, let's limit ourselves
the observed frequencies in the test:
```{r}
astrochron::multiTest(cb(spec_low,c(1,4)))
```


(from astrochron)
if, a priori, you are only concerned with the Milankovitch frequency bands, restrict your analysis to those bands (as constrained by available sedimentation rate estimates and the frequency resolution of the spectrum). in the example below, 
the mtm bandwidth resolution is employed to search frequencies nearby the  Milankovitch-target periods.

```{r}
flow=c((1/400)-0.003,(1/100)-0.003,(1/41)-0.003,(1/20)-0.003)
fhigh=c((1/400)+0.003,(1/100)+0.003,(1/41)+0.003,(1/20)+0.003)
astrochron::multiTest(cb(spec_low,c(1,4)),flow=flow,fhigh=fhigh)
```


But what if you didn't have such a strict prior? It seems to me that we just put the answer in, here.

Now do the same with Bayesian Frequency Selection

```{r}
library(tuts)
# Prepare the data for TUTS. Arbitrarily place uncertainties of 1ka at every point
y=ex[,2]
ti.mu= ex[,1]
ti.sd= rep(1, length(ti.mu))

#Run simulations:
n.sim=1000
WRAP=tuwrap(y=y,ti.mu=ti.mu,ti.sd=ti.sd,n.sim=1000)

#Compare the models:
summary(WRAP)

# Individual summaries, plots and diagnostics:
# Note: Accessing of individual summaries, diagnostics and plots is presented in manuals of all of the funcions contained in the wrapper.SOme examples:

plot(WRAP$BFS,type='periodogram')
summary(WRAP$BFS,tCI=0.99)


```

Oops. That escalated quickly. 



# =====

Try executing this chunk by clicking the *Run* button within the chunk or by placing your cursor inside it and pressing *Cmd+Shift+Enter*. 

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file). 

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.




