---
title: "Ensemble regression and calibration-in-time"
author: "Nick McKay"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ensemble regression and calibration-in-time}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
  
```{r,echo = FALSE}
knitr::opts_chunk$set(fig.retina = 2, fig.width = 8,fig.height = 6) 
knitr::opts_knit$set(progress = FALSE,verbose = FALSE)
```

# Ensemble Regression and Calibration-in-time

Here, we replicate the analysis of Boldt et al. (2015), performing age-uncertain calibration-in-time on a chlorophyll reflectance record from northern Alaska, using geoChronR.

The challenge of age-uncertain calibration-in-time is that age uncertainty affects both the calibration model (the relation between the proxy data and instrumental data) and the reconstruction (the timing of events in the reconstruction). geoChronR simplifies handling these issues. 


Let's start by loading the packages we'll need. 

```{r, results = FALSE, warning = FALSE, message= FALSE }
library(lipdR) #to read and write LiPD files
library(geoChronR) #of course
library(readr) #to load in the instrumetnal data we need
library(ggplot2) #for plotting
``` 

## Load the LiPD file
OK, we'll begin by loading in the Kurupa Lake record from Boldt et al., 2015. `The system.file(...)` part of this pulls the example file from the package directory. You'd like just enter the path as a string for typical use.
```{r}
K <- lipdR::readLipd("http://lipdverse.org/geochronr-examples/Kurupa.Boldt.2015.lpd")
```

## Check out the contents
```{r}
sp <- plotSummary(K,paleo.data.var = "RABD",summary.font.size = 6)
print(sp)
```


## Create an age model with Bacon

```{r,results="hide",fig.keep="all"}
K <- runBacon(K,
              labIDVar = 'labID', 
              age14CVar = 'age14C', 
              age14CuncertaintyVar = 'age14CUncertainty',
              ageVar = 'age', 
              ageUncertaintyVar = 'ageUncertainty', 
              depthVar = 'depth', 
              reservoirAge14CVar = NULL, 
              reservoirAge14CUncertaintyVar = NULL, 
              rejectedAgesVar = NULL,
              baconAccMean = 10,
              baconThick = 7,
              ask = FALSE,
              baconDir = "~/Cores",
              suggest = FALSE)
```

## And plot the ensemble output

```{r}
plotChron(K,age.var = "ageEnsemble",distScale = 0.2)
```


## Map the age ensemble to the paleodata table
This is to get ensemble age estimates for each depth in the paleoData measurement table

```{r}
K <- mapAgeEnsembleToPaleoData(K,age.var = "ageEnsemble")
```


## Select the paleodata age ensemble, and RABD data that we'd like to regress and calibrate

```{r}
kae <-  selectData(K,"ageEnsemble")
rabd <- selectData(K,"RABD")
```


## Now load in the instrumental data we want to correlate and regress agains
```{r}
kurupa.instrumental <- readr::read_csv("http://lipdverse.org/geochronr-examples/KurupaInstrumental.csv")
```

## Check age/time units before proceeding

```{r}
kae$units
```

yep, we need to convert the units from BP to AD

```{r}
kae = convertBP2AD(kae)
```

## Create a "variable list" for the instrumental data

```{r}
kyear = list()
kyear$values = kurupa.instrumental[,1]
kyear$variableName = "year"
kyear$units = "AD"

kinst = list()
kinst$values = kurupa.instrumental[,2]
kinst$variableName = "Temperature"
kinst$units = "deg (C)"
```


## Calculate an ensmeble correlation between the RABD and local summer temperature data
```{r,results="hide",fig.keep="all"}
corout = corEns(kae,rabd,kyear,kinst,binstep=2,percentiles = c(.05,.5,.95 ))
```

## And plot the output

```{r}
plotCorEns(corout)
```

Mixed results. But encouraging enough to move forward.

## Perform ensemble regression
OK, you've convinced yourself that you want to use RABD to model temperature back through time. We can do this simply (perhaps naively) with regession, and lets do it with age uncertainty, both in the building of the model, and the reconstructing
```{r,results="hide",fig.keep="all"}
regout = regressEns(kae,rabd,timeY =kyear,valuesY =kinst,binstep=3,recon.binvec = seq(-4010,2010,by=20),percentiles = c(5,50,95))
```

## And plot the output

```{r}
regPlots = plotRegressEns(regout,alp = 0.01)
```

This result is consistent with that produced by Boldt et al., (2015), and was much simpler to produce with GeoChronR.








```

