---
title: "Filtering, mapping and plotting LiPD objects"
author: "Nick McKay"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Filtering and mapping}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
#LiPD filtering and mappings


```{r, results = FALSE, warning = FALSE, message= FALSE }
library(lipdR)
library(geoChronR)
library(ggplot2)
library(tidyverse)
```


First let's load some data
```{r, results = 'hide', cache=TRUE}
iso <- readLipd("http://lipdverse.org/iso2k/1_0_0/iso2k1_0_0.zip")
```

To quickly map a bunch of LiPD files, use `mapLipd()`

```{r}
mapLipd(iso, global = TRUE,size = 3) + ggtitle("iso2k")
```

Now we've seen an overview of all the files we've loaded, but now we'd like to filter the data.

First, we'll extract a TS object, the primary way to work with multiple LiPD files. A TS is a list object that has an entry for each, column, and is readily queryable. 

```{r}
TS <- extractTs(iso)
```

Now we'll filter it based on a range of criteria. For simple filtering, that include queries like `>` `<` or `==`, you can use `filterTs`

```{r}

iTS <- filterTs(TS, "paleoData_iso2kPrimaryTimeseries == TRUE")

```


For more complex queries, however, it's often more efficient to pull vectors of the metadata that you'd like to query and build and index

```{r}
lat <- pullTsVariable(iTS,"geo_latitude")
lon <- pullTsVariable(iTS,"geo_longitude")
var.name <- pullTsVariable(iTS,"paleoData_variableName")

index <- which(between(lat,10,50) & between (lon,70,170) & var.name == "d18O")

gTS <- iTS[index]
```



OK, now we've filtered down to what we'd like to map. We can use `mapTs()` to map this. 

```{r}
mapTs(gTS)
```

You can change the mapping options by passing options to `basemap()` which creates the base map. Try `?basemap` for options. 



##Time availability

You can also quickly make a plot of availability through time:

```{r}
plotTimeAvailabilityTs(gTS,age.range = c(1,2000),age.var = "year")
```


Finally, if you just want a quick overview, `plotSummaryTs()` will quickly combine these two plots, but with fewer options

```{r}
plotSummaryTs(gTS,age.var = "year", age.range = c(0,2000))
```







